Error Reporting Parsers: a Monad Transformer Approach

### 2013-06-24 article.tex
Parsers:
 - review section showing how transformers helped the code

 - merging code changes into article

 - parser version 3 explanation

With the additional rule that whitespace and comments may appear in any 
amount before any token.  Tokens are:
\begin{itemize}
  \item \verb+{+
  \item \verb+}+
  \item \verb+(+
  \item \verb+)+
  \item \verb+Symbol+
\end{itemize}

@@

Good, it correctly accepts open curly brackets and rejects everything else.
But what about the spec that said that comments and whitespace can occur 
before any token -- will our parser be able to handle that?
\begin{verbatim}
*Main> map (runParser  opencurly) ["}abc", "(abc", "abc"]
Nothing
\end{verbatim}

Of course not -- we haven't told it how to skip comments and whitespace yet!
Let's do that now, first by defining the whitespace and comment patterns, then
by creating a \verb+munch+ combinator that takes in a parser, 
throws away all leading junk, and finally runs the parser.  Also, we rename
the \verb+some+ and \verb+many+ functions from Control.Alternative to more
clearly indicate their semantics:
\begin{verbatim}
many0 = many
many1 = some

whitespace = many1 $ satisfy (flip elem " \n\t\r\f")
comment = pure (:) <*> literal ';' <*> many0 (not1 $ literal '\n')

munch p = many0 (whitespace <+> comment) *> p
\end{verbatim}

@@

And \verb+runParser+ is extended to support the new stack; it assumes we always
start parsing at line 1, column 1:
\begin{verbatim}
runParser :: Parser t a -> [t] -> Either Err (Maybe ((a, [t]), Pos))
runParser p ts = runMaybeT $ runStateT (runStateT p ts) (1, 1)
\end{verbatim}

@@

@@

>>>>

With the additional rule that whitespace and comments may appear in any 
amount before or after tokens.  Tokens are:
\begin{itemize}
  \item \verb+{+
  \item \verb+}+
  \item \verb+(+
  \item \verb+)+
  \item \verb+Symbol+
\end{itemize}

@@

Good, it correctly accepts open curly brackets and rejects everything else.
But what about the spec that said that comments and whitespace can occur 
anywhere before or after tokens -- how should that be handled?

First we need to define the whitespace and comment patterns -- and also rename
the \verb+some+ and \verb+many+ functions from Control.Alternative to more
clearly indicate their semantics:
\begin{verbatim}
many0 = many
many1 = some

whitespace = many1 $ satisfy (flip elem " \n\t\r\f")
comment = pure (:) <*> literal ';' <*> many0 (not1 $ literal '\n')
\end{verbatim}

So what we want to throw away are comments and whitespace; and let's throw them 
away after every token, using a combinator which first parses a token, and then
throws away junk:
\begin{verbatim}
junk = many0 (whitespace <|> comment)

tok p = p <* junk
\end{verbatim}

@@

\verb+runParser+ is extended to support the new stack; it assumes we always
start parsing at line 1, column 1:
\begin{verbatim}
runParser :: Parser t a -> [t] -> Either Err (Maybe ((a, [t]), Pos))
runParser p ts = runMaybeT $ runStateT (runStateT p ts) (1, 1)
\end{verbatim}

@@

We also need to update \verb+item+, such that when it consumes a character, 
it updates the line/column position.  We'll do this by splitting it into two
parts, the part that consumes a single character, and the part that updates the
position:
\begin{verbatim}
basic_item :: (MonadState [t] m, Alternative m) => m t
basic_item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> empty;

item :: Parser Char Char
item = basic_item >>= \x -> updateState (f x) >> pure x
  where f '\n' (ln, c) = (ln + 1, 1)
        f _    (ln, c) = (ln, c + 1)
\end{verbatim}

For simplicity, only '\n' characters will count as newlines; every other character,
including tabs, will count as a single column on the same line.

@@

\section{Monad transformer motivation}
Let's take a quick look at what we've done from a different perspective that
may help motivate why monad transformers were used.

In the first parser, we could parse valid input, but invalid input left us with
absolutely no idea what the problem was or where:
\begin{verbatim}
*Main> runParser woof " {define const {lambda {x x} x}} "
Nothing
\end{verbatim}

The second parser reported what the problem was, but not where:
\begin{verbatim}
*Main> runParser woof " {define const {lambda {x x} x}} "
Left "lambda: duplicate parameter names"
\end{verbatim}

And the third parser was able to report both what and where the problem occurred:
\begin{verbatim}
*Main> runParser woof " {define const {lambda {x x} x}} "
Left ("lambda: duplicate parameter names",(1,25))
\end{verbatim}

Thus, a key to the usefulness of the last parser is that it supports monadic
effects -- backtracking, error reporting, and state.  The parsers which supported 
fewer effects were not as useful.

This emphasizes some major advantages of using monad transformers:
\begin{itemize}
 \item it's easy to create stacks supporting multiple effects
 \item type class instances are taken care of by the library, so stacks already
       support many useful combinators
 \item it's easy to change the composition and order of stacks
\end{itemize}

\Matt{hmm, the last point can be a disadvantage as well}

We benefitted from these advantages when many of the parsers didn't have to change 
at all from one version to the next (although some of their types did change).
